The migration pair birth/death proposal aims to add or remove a pair of migration nodes from the migration history, reassigning demes as necessary to require a consistent migration history. With probability $\frac{1}{2}$ a pair birth proposal is attempted, and otherwise a pair death proposal is attempted.

\subsubsection{Ewing et al. (2004) Version}
	Begin by selecting an edge $e = \langle i,j \rangle \in \calE$ is selected uniformly at random. If a migration pair birth event is attempted, two new nodes $\ihat_1, \ihat_2$ are placed at locations uniformly along $e$, hence splitting edge $e$ into 3 new edges. The demes of these 3 edges are then updated so that the outer edges (ending on either $i$ or $j$) have the same deme as edge $e$, and the central edge $\langle \ihat_1, \ihat_2 \rangle$ has a deme not equal to that of edge $e$.
	
	Otherwise, if a migration pair death event is attempted, the nodes $i$ and $j$ are removed from the tree subject to both $i$ and $j$ being migration nodes, and all of the edges adjacent to $e$ arising from the same deme. If either of these conditions fail, the proposal is assigned a proposal ratio of 0.
	
	The proposal ratio for the migration pair birth proposal is then given by
		\begin{align*}
			Q_B(\calT' | \calT) & = \bbP [\substack{\text{Selected} \\ \text{edge}}] \cdot \bbP [\substack{\text{Selected} \\ \text{deme}}] \cdot \bbP [\substack{\text{New node} \\ \text{times}}] \\
			& = \frac{1}{2n + M - 2} \cdot \frac{1}{d-1} \cdot \frac{2}{\delta t^2} \\
			& = \frac{2}{(2n + M - 2) (d-1) \delta t^2}, \\
			Q_B (\calT | \calT') & = \bbP [ \substack{\text{Selected} \\ \text{edge}}] \\
			& = \frac{1}{2n + M}, \\
			\frac{Q_B (\calT | \calT')}{Q_B (\calT' | \calT)} & = \frac{(2n + M - 2) (d-1) \delta t^2}{2(2n + M)}
		\end{align*}
	where $\delta t$ denotes the time between the targetted pair of migration nodes.
	
	Similarly, the proposal ratio for the migration death proposal is given by
		\[
			\frac{Q_D( \calT | \calT')}{Q_D (\calT' | \calT)} = \frac{2 (2n+M-2)}{(2n + M - 4) (d-1) \delta t^2}.
		\]

\subsubsection{Updated Version 1}
	Begin by selecting a branch of the \textbf{genealogy}, with probability proportional to its length (sufficient to sample a point uniformly along the tree and take the edge of the genealogy where it lands). If a migration pair birth is proposal is attempted, sample two locations uniformly along the selected branch and a deme selected uniformly from the $d$ possible demes. The proposal will fail if
	
	\begin{enumerate}[(i)]
		\item There exists a migration event between the two proposed times
		\item The deme selected leads to an incompatible deme labelling (rejected via the likelihood not the proposal ratio).
	\end{enumerate}
	
	If a migration death proposal is attempted, let $M_b$ denote the number of migration nodes on the selected branch. If $M_b \geq 2$, select a pair of consecutive migration nodes along the branch (if $M_b < 2$, reject the proposal). Then the proposal is completed by deleting the two selected migration nodes.The proposal will fail if
	
	\begin{enumerate}[(i)]
		\item There are fewer than 2 migration nodes along the branch ($M_b < 2$) 
		\item The exterior demes lead to an incompatible deme labelling (rejected via the likelihood not the proposal ratio).
	\end{enumerate}
	
	This change to the proposal mechanism results in the new proposal ratios
	\begin{align*}
		Q_B(\calT' | \calT) & = \bbP[ \substack{\text{Selected} \\ \text{branch}}] \cdot \bbP[ \substack{\text{New node} \\ \text{times}}] \cdot \bbP[\substack{\text{Selected} \\ \text{deme}}] \\
		& = \frac{\calL_b}{\calL} \cdot \frac{2}{\calL_b^2} \cdot \frac{1}{d} \\
		& = \frac{2}{\calL \calL_b d}, \\
		Q_B (\calT | \calT') & = \bbP[\substack{\text{Selected} \\ \text{branch}}] \cdot \bbP[\substack{\text{Selected pair} \\ \text{of nodes}}] \\
		& = \frac{\calL_b}{\calL} \cdot \frac{1}{M_b - 1} \\
		& = \frac{\calL_b}{\calL (M_b - 1)}, \\
		\frac{Q_B (\calT | \calT')}{Q_B (\calT' | \calT)} & = \frac{\calL_b ^2 d}{2(M_b - 1)}
	\end{align*}

\subsubsection{Updated Version 2 (Jere's version?)}
	Pair birth:
	\begin{enumerate}[(i)]
		\item Sample branch from the genealogy proportional to its length
		\item Sample two locations uniformly along the branch
		\item Sample 2 demes to be pulled inwards from the proposed locations (can sample deme uniformly from set of demes not equal to current deme at that location)
		\item Pull proposed demes inwards from proposed locations towards centre of branch until hitting a migration node (or the other sampled location)
	\end{enumerate}
	
	\begin{align*}
		Q_B (\calT' | \calT) & = \bbP[\substack{\text{Selected} \\ \text{branch}}] \cdot \bbP[\substack{\text{New node} \\ \text{times}}] \cdot \bbP[\substack{\text{Selected} \\ \text{demes}}] \\
		& = \frac{\calL_b}{\calL} \cdot \frac{2}{\calL_b^2} \cdot \frac{1}{(d-1)^2}.
	\end{align*}
	
	\begin{align*}
		\frac{Q_D(\calT | \calT')}{Q_B(\calT' | \calT)} & = \frac{\frac{\calL_b}{\calL} \cdot \frac{2}{(M_b + 1)(M_b + 2)} \cdot \frac{1}{(d-1)^2}}{\frac{\calL_b}{\calL} \cdot \frac{2}{\calL_b^2} \cdot \frac{1}{(d-1)^2}} \\
		& = \frac{\calL_b^2}{(M_b + 1) (M_b + 2)}
	\end{align*}
	\\[2ex]
	
	Pair death:
	\begin{enumerate}[(i)]
		\item Sample branch from the genealogy proportional to its length
		\item If there are at least 2 migration events on the branch, sample two migration nodes
		\item Sample 2 demes to be pulled inwards from the proposed locations (if the locations are consecutive, special case recovers Ewing pair death move if both proposed demes are the same)
		\item Pull proposed demes inwards from proposed locations towards centre of branch until hitting a migration node (including the other sampled migration node)
	\end{enumerate}
	\begin{align*}
		Q_D (\calT' | \calT) & = \bbP[\substack{\text{Selected} \\ \text{branch}}] \cdot \bbP[\substack{\text{Selected} \\ \text{nodes}}] \cdot \bbP[\substack{\text{Selected} \\ \text{demes}}] \\
		& = \frac{\calL_b}{\calL} \cdot \frac{2}{M_b (M_b - 1)} \cdot \frac{1}{(d-1)^2}
	\end{align*}
	
	\begin{align*}
		\frac{Q_B(\calT | \calT')}{Q_D (\calT' | \calT)} & = \frac{\frac{\calL_b}{\calL} \cdot \frac{2}{\calL_b^2} \cdot \frac{1}{(d-1)^2}}{\frac{\calL_b}{\calL} \cdot \frac{2}{M_b (M_b - 1)} \cdot \frac{1}{(d-1)^2}} \\
		& = \frac{M_b (M_b - 1)}{\calL_b^2}.
	\end{align*}