---
title: "Code Testing"
author: "Ian Roberts"
date: "09/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
library(ape)
```

\newcommand{\calJ}{\mathcal{J}}
\newcommand{\calT}{\mathcal{T}}

## Structured Simulation Testing
Generate datasets of size $n$ observations with

  * Random sample times uniform between two years (default 2010 to 2021)
  * Random demes sampled uniformly at random from 1:n.deme (default n.deme = 2)

```{r}
Test.data <- function(n, n.deme = 2, lower.year = 2010, upper.year = 2021){
  data <- matrix(0,nrow = n, ncol = 3)
  data[,1] <- 1:n
  data[,2] <- runif(n, min = lower.year, max = upper.year + 1)
  data[,3] <- sample.int(n.deme, n, replace = TRUE)
  data
}
```

Run $N$ repetitions computing the recursive likelihood under Structured.sim and the likelihood from structured.likelihood with unique test datasets.
```{r}
N <- 100  #Number of repetitions
n <- 100  #Sample size
n.deme <- 3
migration.mat <- matrix(runif(n.deme^2, 0, 0.1),n.deme, n.deme)

likelihoods <- matrix(0,nrow = N,ncol = 2)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
```
Consistently returns discrepancies between calculated likelihoods of order $10^{-13}$ suggesting accurate computations are being made.

*Initially a discrepancy arose from removing the diagonal of migration.mat in Structured.sim but not in structured.likelihood*

Repeating the testing with a greater number of demes:
```{r}
N <- 100
n.deme <- 5

migration.mat <- matrix(runif(n.deme^2),n.deme, n.deme)

likelihoods <- matrix(0,nrow = N,ncol = 2)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
```

confirm that the likelihoods computed under structured.sim agree with the heterochronous coalescent without migration when there is a single deme.
```{r}
n.deme <- 1
N <- 100

migration.mat <- matrix(1,1,1)

likelihoods <- matrix(0,N,3)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
  likelihoods[i,3] <- phylo.likelihood(phylo,1,1)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
max(abs(likelihoods[,2] - likelihoods[,3]))
```

Consistently, the likelihoods calculated for the structured coalescent with a single deme (using structured.likelihood) correspond exactly to the heterochronous coalescent (using phylo.likelihood)


## MCMC Implementation Testing

Begin by testing a MCMC algorithm with a single type of move (migration birth/death).

TESTING MIGRATION BIRTH/DEATH MOVE ONLY

```{r}
n <- 10
n.deme <- 5
migration.mat <- matrix(runif(n.deme^2, 0, 0.1),n.deme, n.deme)
phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
structured.plot(phylo); nodelabels(frame = "none"); tiplabels(frame = "none")

ED <- phylo.to.ed(phylo)

for (i in 1 : 10){
  Out <- ed.mig.death.new(ED, n.deme); Out$prop.ratio
  if (Out$prop.ratio > 0){
    new.phylo <- ed.to.phylo(Out$ED)
    structured.plot(new.phylo); nodelabels(frame = "none"); tiplabels(frame = "none")
  }
}


```

Burn in phase 100 iterations, main run 1000 iterations; prior ratio 1 and likelihood ratio given by ratio of poisson(1) distributions on number of migration events. Select migration birth w/prob p and migration death w/prob 1-p

```{r}
set.seed(1)
N0 <-  10^2 #Burn in phase
N <- 10^3
p <- 1/2
sample <- numeric(N)

n <- 100
n.deme <- 3
migration.mat <- matrix(runif(n.deme^2, 0, 1),n.deme, n.deme)
phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)  #Initial tree
ED <- phylo.to.ed(phylo)

for (i in 1:(N0 + N)){
  U <- runif(1)
  
  if (U <= p){ #Birth
    proposal <- ed.mig.birth.3(ED, n.deme, FALSE) #ed.mig.birth.new(ED, n.deme)
  } else{ #Death
    proposal <- ed.mig.death.3(ED, n.deme, FALSE) #ed.mig.death.new(ED, n.deme)
  }
  
  ED.n.mig <- length(ED[is.na(ED[,4]) & (! is.na(ED[,3])), 1])  #Number of migrations in current tree
  prop.ED <- proposal$ED
  prop.n.mig <- length(prop.ED[is.na(prop.ED[,4]) & (! is.na(prop.ED[,3])), 1])  #Number of migrations in proposal
  
  accept.prob <- min(1, factorial(ED.n.mig)/factorial(prop.n.mig) * proposal$prop.ratio)
  
  V <- runif(1)
  
  if (V <= accept.prob){
    ED <- proposal$ED
  }
  
  if (i > N0){
      sample[i-N0] <- length(ED[is.na(ED[,4]) & (! is.na(ED[,3])), 1])
  }
}
summary(as.factor(sample))
structured.plot(ed.to.phylo(ED)) #;nodelabels(frame = "none")
structured.plot(phylo) #;nodelabels(frame = "none")
hist(sample)
```


Construct a tree with $n$ homochronous leaves such that all lineages migrate into deme 1 almost immediately after the sample is drawn and no more migrations occur. Assume also that all generated data do not begin in deme 1.

Initialising at this tree reduces the possiblity of having to reject any proposal due to a leaf being included in the subtree and also starts at an almost minimal number of migration events.

```{r}
n <- 20
n.deme <- 3
DATA <- matrix(c(1:n, rep(0,n), sample(2:n.deme, n, replace = TRUE)), n)

migration.matrix <- matrix(0.3, n.deme,n.deme)
diag(migration.matrix) <- 0

tree.top <- Homochronous.Sim(1:n, 1, 1)
tree.top$node.deme <- rep(1, 2*n-1)
fake.ED <- phylo.to.ed(tree.top)

max.node <- max(fake.ED[,1])
node.labels <- fake.ED[,1:4]

for (i in 1:n){
  node.labels[which(node.labels == i)] <- max.node + i
}

fake.ED[,1:4] <- node.labels

for (i in 1:n){
  fake.ED <- rbind(fake.ED, c(i, max.node + i, NA, NA, sample((1:n.deme)[-fake.ED[i, 5]], 1), fake.ED[i, 6] + 1e-3))
  fake.ED[i, 3] <- i
}

structured.plot(ed.to.phylo(fake.ED))
```
Run the MCMC with Poisson likelihood for this artificial tree. The minimum number of migration events possible in this tree is 20 as none of the current migration nodes may be removed. (Should obtain a 20 + Poi(1) sample???)

```{r}
N <- 1e3
p <- 0.5
sample <- numeric(N)

ED <- fake.ED
for (i in 1:N){
  if (i/2 != floor(i/2)){ #Birth
    proposal <- ed.mig.birth.3(ED, n.deme) #ed.mig.birth.new(ED, n.deme)
  } else{ #Death
    proposal <- ed.mig.death.3(ED, n.deme) #ed.mig.death.new(ED, n.deme)
  }
  
  ED.n.mig <- length(ED[is.na(ED[,4]) & (! is.na(ED[,3])), 1])  #Number of migrations in current tree
  prop.ED <- proposal$ED
  prop.n.mig <- length(prop.ED[is.na(prop.ED[,4]) & (! is.na(prop.ED[,3])), 1])  #Number of migrations in proposal
  
  accept.prob <- min(1, factorial(ED.n.mig)/factorial(prop.n.mig) * proposal$prop.ratio)
  
  V <- runif(1)
  
  if (V <= accept.prob){
    ED <- proposal$ED
  }
  
  sample[i] <- length(ED[is.na(ED[,4]) & (! is.na(ED[,3])), 1])
}
structured.plot(ed.to.phylo(fake.ED))
structured.plot(ed.to.phylo(ED))
hist(sample)
structured.plot(ed.to.phylo(proposal$ED))
```



## Testing MCMC with a single-colour input

Generate a structured coalescent genealogy with no migration events and run the MCMC algorithm with Poisson likelihoods. Still permit migration events to be added, but also permit leaf demes to be changed as necessary.

```{r}
#set.seed(10)
N <- 1e5  #Number of MCMC iterations

n <- 10  #Number of leaves in the tree
n.deme <- 5
phylo <- Homochronous.Sim(1:n, 1, 1)
phylo$node.deme <- rep(1, 2*n - 1)

ED <- phylo.to.ed(phylo)
structured.plot(phylo)

migration.mat <- matrix(0.3, n.deme, n.deme)
diag(migration.mat) <- 0

Sample <- numeric(N)

for (i in 1:N){
  #if (i >= 118){ browser()}
  
  U <- runif(1)
  
  if (U <= 0.5){
    proposal <- ed.mig.birth.3(ED, n.deme, FALSE)
  } else {
    proposal <- ed.mig.death.3(ED, n.deme, FALSE)
  }
  
  ED.n.mig <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])
  prop.n.mig <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
  
  accept.prob <- min(1, factorial(ED.n.mig)/factorial(prop.n.mig) * proposal$prop.ratio)
  
  V <- runif(1)
  
  if (V <= accept.prob){  #ACCEPT
    ED <- proposal$ED
    #browser()
    #structured.plot(ed.to.phylo(ED))
    Sample[i] <- prop.n.mig
  } else{  #REJECT
    Sample[i] <- ED.n.mig
  }
}

hist(Sample, breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, 1)))
```

## Testing MCMC algorithm with multi-colour input tree

Stick with homochronous leaves, but relax the assumption that the initial tree remains in deme 1 throughout. Generate a tree using Structured.sim with $n=10$ tips and initial demes sampled from 1:n.deme with replacement. Add also 100 burn-in iterations to avoid skewing the number of migrations by the intial tree.

```{r}
N0 <- 1e2  #Number of burn-in iterations
N <- 1e5  #Number of MCMC iterations

n <- 10  #Number of leaves in the tree
n.deme <- 5

Data <- matrix(c(1:n, rep(0, n), sample.vector(1:n.deme, n, TRUE)), n, 3)
migration.mat <- matrix(0.3, n.deme, n.deme)
diag(migration.mat) <- 0

phylo <- Structured.sim(Data, 1, 1, n.deme, migration.mat)

ED <- phylo.to.ed(phylo)

Sample <- numeric(N)

for (i in (-N0):N){
  #if (i >= 118){ browser()}
  
  U <- runif(1)
  
  if (U <= 0.5){
    proposal <- ed.mig.birth.3(ED, n.deme, FALSE)
  } else {
    proposal <- ed.mig.death.3(ED, n.deme, FALSE)
  }
  
  ED.n.mig <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])
  prop.n.mig <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
  
  accept.prob <- min(1, factorial(ED.n.mig)/factorial(prop.n.mig) * proposal$prop.ratio)
  
  V <- runif(1)
  
  if (V <= accept.prob){  #ACCEPT
    ED <- proposal$ED
    #browser()
    #structured.plot(ed.to.phylo(ED))
    if (i > 0){
      Sample[i] <- prop.n.mig
    }
  } else{  #REJECT
    if (i > 0){
      Sample[i] <- ED.n.mig
    }
  }
}

hist(Sample, breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, 1)))

```

## Testing MCMC with higher rate Poisson likelihood

In place of the Poisson(1) likelihood used so far in the MCMC schenes, instead use a Poisson($\lambda$). Begin with $\lambda = 10$.

```{r}
lambda <- 10

N0 <- 0 #1e2  #Number of burn-in iterations
N <- 1e5  #Number of MCMC iterations

p <- 1/2 #Probability of proposing a migration birth

n <- 10  #Number of leaves in the tree
n.deme <- 10

Data <- matrix(c(1:n, rep(0, n), sample.vector(1:n.deme, n, TRUE)), n, 3)
migration.mat <- matrix(0.03, n.deme, n.deme)
diag(migration.mat) <- 0

phylo <- Structured.sim(Data, 1, 1, n.deme, migration.mat)

ED <- phylo.to.ed(phylo)

Sample <- numeric(N)

for (i in (-N0):N){
  U <- runif(1)
  M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
  if (U <= p){
    proposal <- ed.mig.birth.4(ED, n.deme, FALSE)
    accept.prob <- min(1, lambda/(M+1))
    
  } else {
    proposal <- ed.mig.death.4(ED, n.deme, FALSE)
    accept.prob <- min(1, M/lambda)
    
  }
  
  V <- runif(1)
  
  if (V <= accept.prob){  #ACCEPT
    ED <- proposal$ED
    if (i > 0){
      Sample[i] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
      }
  } else{  #REJECT
    if (i > 0){
      Sample[i] <- M
    }
  }
}

hist(Sample, breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
lines(0:150, dpois(0:150, lambda))
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, lambda)))
plot(Sample, type = 'l')
```



## Migration Pair Birth/Death MCMC

Now test the Poisson($\lambda$) MCMC algorithm with only the use of migration pair birth/death moves.

```{r}
set.seed(10)
lambda <- 10

N0 <- 1e3 #Number of burn-in iterations
N <- 1e5  #Number of MCMC iterations

p <- 1/2 #Probability of proposing a migration birth

n <- 10  #Number of leaves in the tree
n.deme <- 2

Data <- matrix(c(1:n, rep(0, n), sample.vector(1:n.deme, n, TRUE)), n, 3)
migration.mat <- matrix(0.03, n.deme, n.deme)
diag(migration.mat) <- 0

phylo <- Homochronous.Sim(1:n, 1, 1)
phylo$node.deme <- rep(1, 2 * n - 1)

ED <- phylo.to.ed(phylo)

Sample <- numeric(N)

##### CALCULATE TREE LENGTH
  root.node <- ED[is.na(ED[,2]), 1]
  
  edge.length <- numeric(dim(ED)[1])
  non.root.nodes <- ED[ED[,1] != root.node, 1]
  non.root.nodes <- non.root.nodes[!is.na(non.root.nodes)]
  for (j in non.root.nodes){
    node.row <- which(ED[,1] == j)
    parent.row <- which(ED[,1] == ED[node.row, 2])
    edge.length[node.row] <- ED[node.row, 6] - ED[parent.row, 6]
  }
  tree.length <- sum(edge.length)  #Total tree length


for (i in (-N0):N){
  #if (i >= 1170 ){
  #  browser()
  #  print(ED)
  #  }
  U <- runif(1)
  M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
  if (U <= p){
    proposal <- ed.mig.pair.birth(ED, n.deme)
    accept.prob <- min(1, (lambda / ((n.deme - 1) * tree.length))^2 * proposal$prop.ratio )
    
  } else {
    proposal <- ed.mig.pair.death(ED, n.deme)
    accept.prob <- min(1, ((n.deme - 1) * tree.length / lambda)^2 * proposal$prop.ratio)
  }
  
  V <- runif(1)
  
  if (V <= accept.prob){  #ACCEPT
    ED <- proposal$ED
    if (i > 0){
      Sample[i] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
      }
  } else{  #REJECT
    if (i > 0){
      Sample[i] <- M
    }
  }
}

hist(Sample, breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
k <- seq(0, 150, 2)
lines(k, 2*lambda^k * exp(-lambda) / ((1 + exp(-2 * lambda)) * factorial(k)))
#lines(0:150, dpois(0:150, lambda/2))
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, lambda)))
plot(Sample, type = 'l')
```

```{r}
n <- 3
n.deme <- 5

Data <- matrix(c(1:n, rep(0, n), sample.vector(1:n.deme, n, TRUE)), n, 3)
migration.mat <- matrix(0.03, n.deme, n.deme)
diag(migration.mat) <- 0

phylo <- Structured.sim(Data, 1, 1, n.deme, migration.mat)
structured.plot(phylo); tiplabels(frame = "none"); nodelabels(frame = "none")
ED <- phylo.to.ed(phylo)

for (i in 1:100){
  ED <- ed.mig.pair.birth.2(ED, n.deme)$ED
}
structured.plot(ed.to.phylo(ED))
for (i in 1:100){
  ED <- ed.mig.pair.death.2(ED, n.deme)$ED
}
```


```{r}
#set.seed(10)
lambda <- 50

N0 <- 1e3 #Number of burn-in iterations
N <- 1e7  #Number of MCMC iterations

p <- 1/2 #Probability of proposing a migration birth

n <- 20  #Number of leaves in the tree
n.deme <- 10

phylo <- Homochronous.Sim(1:n, 1, 1)
phylo$node.deme <- rep(1, 2 * n - 1)

ED <- phylo.to.ed(phylo)

Sample <- matrix(0, N, 3)

##### CALCULATE TREE LENGTH
  root.node <- ED[is.na(ED[,2]), 1]
  
  edge.length <- numeric(dim(ED)[1])
  non.root.nodes <- ED[ED[,1] != root.node, 1]
  non.root.nodes <- non.root.nodes[!is.na(non.root.nodes)]
  for (j in non.root.nodes){
    node.row <- which(ED[,1] == j)
    parent.row <- which(ED[,1] == ED[node.row, 2])
    edge.length[node.row] <- ED[node.row, 6] - ED[parent.row, 6]
  }
  tree.length <- sum(edge.length)  #Total tree length

for (i in (-N0):N){
  #if (i >= 1170 ){
  #  browser()
  #  print(ED)
  #  }
  
  W <- runif(1)

  if (W < 0.01){
    U <- runif(1)
    M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
    if (U <= p){
      proposal <- ed.mig.birth.4(ED, n.deme, FALSE)
      accept.prob <- min(1, lambda/(M+1))
      if (i > 0){
        Sample[i, 2] <- 1
      }
      
    } else {
      proposal <- ed.mig.death.4(ED, n.deme, FALSE)
      accept.prob <- min(1, M/lambda)
      if (i > 0){
        Sample[i, 2] <- 2
      }
    }
    
    V <- runif(1)
    
    if (V <= accept.prob){  #ACCEPT
      #browser()
      ED <- proposal$ED
      if (i > 0){
        Sample[i,1] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
        Sample[i,3] <- 1
        }
    } else{  #REJECT
      if (i > 0){
        Sample[i,1] <- M
      }
    }
  } else{
    U <- runif(1)
    M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
    if (U <= p){
      proposal <- ed.mig.pair.birth(ED, n.deme)
      accept.prob <- min(1, (lambda / ((n.deme - 1) * tree.length))^2 * proposal$prop.ratio )
      if (i > 0){
        Sample[i, 2] <- 3
      }
    } else {
      proposal <- ed.mig.pair.death(ED, n.deme)
      accept.prob <- min(1, ((n.deme - 1) * tree.length / lambda)^2 * proposal$prop.ratio)
      if (i > 0){
        Sample[i, 2] <- 4
      }
    }
    
    V <- runif(1)
    
    if (V <= accept.prob){  #ACCEPT
      ED <- proposal$ED
      if (i > 0){
        Sample[i,1] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
        Sample[i,3] <- 1
        }
    } else{  #REJECT
      if (i > 0){
        Sample[i,1] <- M
      }
    }
  }
}
  
hist(Sample[,1], breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
k <- seq(0, 150, 2)
#lines(k, 2*lambda^k * exp(-lambda) / ((1 + exp(-2 * lambda)) * factorial(k)))
lines(0:150, dpois(0:150, lambda))
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, lambda)))
plot(Sample[,1], type = 'l')
summary(as.factor(Sample[,2]))
sum(Sample[Sample[,2] %in% c(3, 4),3] )
sum(Sample[,3])

```


## Full MCMC Scheme Testing

```{r}
#set.seed(10)
lambda <- 10

N0 <- 1e3 #Number of burn-in iterations
N <- 1e7  #Number of MCMC iterations

p <- 1/2 #Probability of proposing a migration birth

n <- 3  #Number of leaves in the tree
n.deme <- 3

phylo <- Homochronous.Sim(1:n, 1, 1)
phylo$node.deme <- rep(1, 2 * n - 1)

ED <- phylo.to.ed(phylo)

Sample <- matrix(0, N, 3)

##### CALCULATE TREE LENGTH
  root.node <- ED[is.na(ED[,2]), 1]
  
  edge.length <- numeric(dim(ED)[1])
  non.root.nodes <- ED[ED[,1] != root.node, 1]
  non.root.nodes <- non.root.nodes[!is.na(non.root.nodes)]
  for (j in non.root.nodes){
    node.row <- which(ED[,1] == j)
    parent.row <- which(ED[,1] == ED[node.row, 2])
    edge.length[node.row] <- ED[node.row, 6] - ED[parent.row, 6]
  }
  tree.length <- sum(edge.length)  #Total tree length

for (i in (-N0):N){
  #if (max(ED[,1]) >= 65 ){browser(); print(ED)}
  #if (i == -740) {browser()}
  
  W <- runif(1)

  if (W < 0.01){
    U <- runif(1)
    M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
    if (U <= p){
      proposal <- ed.mig.birth.4(ED, n.deme, FALSE)
      accept.prob <- min(1, lambda/(M+1))
      if (i > 0){
        Sample[i, 2] <- 1
      }
      
    } else {
      proposal <- ed.mig.death.4(ED, n.deme, FALSE)
      accept.prob <- min(1, M/lambda)
      if (i > 0){
        Sample[i, 2] <- 2
      }
    }
    
    V <- runif(1)
    
    if (V <= accept.prob){  #ACCEPT
      #browser()
      ED <- proposal$ED
      if (i > 0){
        Sample[i,1] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
        Sample[i,3] <- 1
        }
    } else{  #REJECT
      if (i > 0){
        Sample[i,1] <- M
      }
    }
  } else if (W < 1/2) {
    U <- runif(1)
    M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
    if (U <= p){
      proposal <- ed.mig.pair.birth(ED, n.deme)
      accept.prob <- min(1, (lambda / ((n.deme - 1) * tree.length))^2 * proposal$prop.ratio )
      if (i > 0){
        Sample[i, 2] <- 3
      }
    } else {
      proposal <- ed.mig.pair.death(ED, n.deme)
      accept.prob <- min(1, ((n.deme - 1) * tree.length / lambda)^2 * proposal$prop.ratio)
      if (i > 0){
        Sample[i, 2] <- 4
      }
    }
    
    V <- runif(1)
    
    if (V <= accept.prob){  #ACCEPT
      ED <- proposal$ED
      if (i > 0){
        Sample[i,1] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
        Sample[i,3] <- 1
        }
    } else{  #REJECT
      if (i > 0){
        Sample[i,1] <- M
      }
    }
  } else{
    U <- runif(1)
    M <- length(ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1])  #Number of migration events in current ED
    if (U <= p){
      proposal <- ed.coal.split(ED, n.deme)
      accept.prob <- min(1, (lambda / ((n.deme - 1) * tree.length)) * proposal$prop.ratio )
      if (i > 0){
        Sample[i, 2] <- 5
      }
    } else {
      proposal <- ed.coal.merge(ED, n.deme)
      accept.prob <- min(1, ((n.deme - 1) * tree.length / lambda) * proposal$prop.ratio)
      if (i > 0){
        Sample[i, 2] <- 6
      }
    }
    
    V <- runif(1)
    
    if (V <= accept.prob){  #ACCEPT
      ED <- proposal$ED
      if (i > 0){
        Sample[i,1] <- length(proposal$ED[ is.na(proposal$ED[,4]) & (!is.na(proposal$ED[,3])), 1])
        Sample[i,3] <- 1
        }
    } else{  #REJECT
      if (i > 0){
        Sample[i,1] <- M
      }
    }
  }
}
  
hist(Sample[,1], breaks = (min(Sample)-0.5):(max(Sample)+1), probability = TRUE)
k <- seq(0, 150, 2)
#lines(k, 2*lambda^k * exp(-lambda) / ((1 + exp(-2 * lambda)) * factorial(k)))
lines(0:150, dpois(0:150, lambda))
print("MCMC Sample"); summary(as.factor(Sample))
print("Poisson Sample"); summary(as.factor(rpois(N, lambda)))
plot(Sample[,1], type = 'l')
summary(as.factor(Sample[,2]))
sum(Sample[Sample[,2] %in% c(3, 4),3] )
sum(Sample[,3])

```
