---
title: "Code Testing"
author: "Ian Roberts"
date: "09/06/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
library(ape)
```

## Structured Simulation Testing
Generate datasets of size $n$ observations with

  * Random sample times uniform between two years (default 2010 to 2021)
  * Random demes sampled uniformly at random from 1:n.deme (default n.deme = 2)

```{r}
Test.data <- function(n, n.deme = 2, lower.year = 2010, upper.year = 2021){
  data <- matrix(0,nrow = n, ncol = 3)
  data[,1] <- 1:n
  data[,2] <- runif(n, min = lower.year, max = upper.year + 1)
  data[,3] <- sample.int(n.deme, n, replace = TRUE)
  data
}
```

Run $N$ repetitions computing the recursive likelihood under Structured.sim and the likelihood from structured.likelihood with unique test datasets.
```{r}
N <- 100  #Number of repetitions
n <- 100  #Sample size
n.deme <- 2
migration.mat <- matrix(runif(n.deme^2),n.deme, n.deme)

likelihoods <- matrix(0,nrow = N,ncol = 2)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
```
Consistently returns discrepancies between calculated likelihoods of order $10^{-13}$ suggesting accurate computations are being made.

*Initially a discrepancy arose from removing the diagonal of migration.mat in Structured.sim but not in structured.likelihood*

Repeating the testing with a greater number of demes:
```{r}
N <- 100
n.deme <- 5

migration.mat <- matrix(runif(n.deme^2),n.deme, n.deme)

likelihoods <- matrix(0,nrow = N,ncol = 2)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
```

confirm that the likelihoods computed under structured.sim agree with the heterochronous coalescent without migration when there is a single deme.
```{r}
n.deme <- 1
N <- 100

migration.mat <- matrix(1,1,1)

likelihoods <- matrix(0,N,3)

for (i in 1:N){
  phylo <- Structured.sim(Test.data(n,n.deme), 1,1,n.deme,migration.mat,FALSE)
  likelihoods[i,1] <- phylo$log.likelihood
  likelihoods[i,2] <- structured.likelihood(phylo,1,1,migration.mat)[1]
  likelihoods[i,3] <- phylo.likelihood(phylo,1,1)[1]
}

max(abs(likelihoods[,1] - likelihoods[,2]))
max(abs(likelihoods[,2] - likelihoods[,3]))
```

Consistently, the likelihoods calculated for the structured coalescent with a single deme (using structured.likelihood) correspond exactly to the heterochronous coalescent (using phylo.likelihood)
