---
title: "SimulationVerification"
author: "Ian Roberts"
date: "22/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
devtools::load_all()
library(ape)
```

## Verifying Structured.sim structured coalescent simulations

Can obtain MLE estimates for the effective population size and migration matrix (with 3 different structures for the migration matrix). Then a sample of size $N$ can be drawn for the MLEs by generating $N$ structured coalescent trees with fixed tip data.

```{r}
### Fixed tip data to be used in all simulations
set.seed(10)
n <- 10
n.deme <- 3

data <- matrix(0,nrow = n, ncol = 3)
data[,1] <- 1:n
data[,2] <- rep(2022, n) #runif(n, min = 2010, max = 2022)
data[,3] <- sample.int(n.deme, n, replace = TRUE)

N <- 1e5 #Sample size
```

### Effective population sizes

MLE for effective population sizes
  \[
    \hat{\theta}_i = \frac{1}{c_i} \sum_r \binom{k_{ir}}{2} \delta t_r
  \]
  
```{r}
#Simulation parameters
gen.length <- 1
effective.pop <- sample(5:25/10, n.deme, replace = TRUE)
migration.matrix <- matrix(1/2, n.deme, n.deme) #matrix(rgamma(n.deme^2, shape = mig.mat.alpha, scale = 1/mig.mat.beta), n.deme, n.deme)
diag(migration.matrix) <- 0

MLE.sample <- matrix(NA, N, n.deme)

for (a in 1 : N){
  phylo <- Structured.sim(data, effective.pop, gen.length, n.deme, migration.matrix, FALSE)
  ED <- phylo.to.ed(phylo)
  
  node.heights <- ED[,6]
  event.times <- sort(unique(node.heights))  #Times of events, root at time=0
  time.increments <- diff(event.times)
  
  root.node <- ED[is.na(ED[,2]), 1]
  coalescence.nodes <- ED[!is.na(ED[,4]),1]
  migration.nodes <- ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1]
  leaf.nodes <- ED[(is.na(ED[,3])) & (is.na(ED[,4])), 1]
  
  k <- matrix(0, nrow = length(event.times) - 1, ncol = n.deme)
  root.row <- which(ED[,1] == root.node)
  root.child.1 <- ED[root.row, 3]
  k[1,ED[which(ED[,1] == root.child.1),5]] <- 2
  for (i in 2 : (length(event.times) - 1)){
    current.rows <- which(ED[,6] == event.times[i])
    k[i,] <- k[i-1,]
    if (length(current.rows) > 1){ #Multiple leaves added simultaneously
      for (j in current.rows){
        k[i, ED[j, 5]] <- k[i, ED[j, 5]] + 1
      }
    } else{
      if (ED[current.rows, 1] %in% migration.nodes){ #Migration event
        k[i, ED[current.rows, 5]] <- k[i, ED[current.rows, 5]] - 1
        current.child <- ED[current.rows, 3]
        current.child.row <- which(ED[,1] == current.child)
        k[i, ED[current.child.row, 5]] <- k[i, ED[current.child.row, 5]] + 1
      } else if (current.rows %in% coalescence.nodes){ #Coalescence event
        k[i, ED[current.rows, 5]] <- k[i, ED[current.rows, 5]] + 1
      } else{ #Single leaf added
        k[i, ED[current.rows, 5]] <- k[i, ED[current.rows, 5]] - 1
      }
    }
  }
  
  C <- ed.node.count(ED, n.deme)$c
  rate.constants <- t(choose(k,2)) %*% time.increments
  
  MLE.sample[a,] <- rate.constants/C
}

upper <- ceiling(50*max(MLE.sample[is.finite(MLE.sample)]))/50
for (b in 1 : n.deme){
  hist(MLE.sample[is.finite(MLE.sample[,b]), b], breaks = seq(0,upper,1/50), freq = FALSE, main = paste("True parameter = ", effective.pop[b]))
  abline(v = mean(MLE.sample[is.finite(MLE.sample[,b]), b]), col = "red", lwd = 2, lty = 2)
  abline(v = effective.pop[b], col = "blue", lwd = 2, lty = 2)
  legend("topright", legend = c("Sample Mean", "Simulation Parameter"), col = c("red", "blue"), lty = 2)
  lines(seq(0,5,0.01), dnorm(seq(0,5,0.01), mean(MLE.sample[is.finite(MLE.sample[,b]), b]), sd(MLE.sample[is.finite(MLE.sample[,b]), b])))
}
```

### ARD (All Rates Different) migration matrix 

MLE for ARD migration matrix
  \[
    \hat{\lambda}_{ij}^{(ARD)} = \frac{m_{ij}}{\sum_r k_{ir} \delta t_r}
  \]

### SR (Symmetric rates) migration matrix

MLE for SR migration matrix
  \[
    \hat{\lambda}_{ij}^{(SR)} = \frac{m_{ij} + m_{ji}}{\sum_r (k_{ir} + k_{jr}) \delta t_r}
  \]

### ER (Equal rates) migration matrix

MLE for ER migration matrix
  \[
    \hat{\lambda} = \frac{M}{(d-1)\mathcal{L}}
  \]

```{r}
#Simulation parameters
gen.length <- 1
effective.pop <- rep(1, n.deme)  #1 / rgamma(n.deme, shape = eff.pop.alpha, scale = 1 /eff.pop.beta)
migration.matrix <- matrix(1/2, n.deme, n.deme) #matrix(rgamma(n.deme^2, shape = mig.mat.alpha, scale = 1/mig.mat.beta), n.deme, n.deme)
diag(migration.matrix) <- 0

MLE.sample <- numeric(N)

for (a in 1:N){
  phylo <- Structured.sim(data, effective.pop, gen.length, n.deme, migration.matrix, FALSE)
  ED <- phylo.to.ed(phylo)
  
  root.row <- which(is.na(ED[,2]))
  tree.length <- 0
  
  for (i in (1 : dim(ED)[1])[-root.row]){
    parent.row <- which(ED[,1] == ED[i,2])
    tree.length <- tree.length + ED[i, 6] - ED[parent.row, 6]
  }
  
  migration.nodes <- ED[ is.na(ED[,4]) & (!is.na(ED[,3])) ,1]
  M <- length(migration.nodes)
  
  MLE.sample[a] <- M / ((n.deme-1) * tree.length)
}
#plot(MLE.sample)
upper <- ceiling(50*max(MLE.sample))/50
hist(MLE.sample, freq = FALSE, breaks = seq(0,upper,1/50))
abline(v = mean(MLE.sample), col = "red", lwd = 2, lty = 2)
lines(seq(0,5,0.01), dnorm(seq(0,5,0.01), mean(MLE.sample), sd(MLE.sample)))
mean(MLE.sample)


```
